---
title: "Building IMDb Top 250 Movies Database"
author: "Mackenzie Ross"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Load Libraries**
```{r}
library(tidyverse)
library(dplyr)
library(RSQLite)
library(DBI)
```

# Load and Clean Data

## Read in IMDb Top 250 Data
```{r}
data <- read.csv('imdb_movies.csv')
```

## Preview Data
```{r}
head(data)
```
The column 'X' is not needed to build the database. The columns 'imbd_votes' and 'imbd_rating' have the b and d inverted, so those column titles need to be changed

### Remove column 'X' and Change Mispelled Column Titles
```{r}
imdb250 <- data %>% select(-c(X)) %>% rename(imdb_votes = imbd_votes, imdb_rating = imbd_rating)
head(imdb250)
```

There is an extra comma at the end of the list of writers in the third row and a writer named "Writers". We will use the gsub() function to remove any potential commas at the end of any of columns that have lists. We will also remove commas from the imdb_votes column and change it to a numeric column.

```{r}
# remove trailing commas
imdb250$genre <- gsub(",$", "", imdb250$genre)
imdb250$cast_id <- gsub(",$", "", imdb250$cast_id)
imdb250$cast_name <- gsub(",$", "", imdb250$cast_name)
imdb250$director_id <- gsub(",$", "", imdb250$director_id )
imdb250$director_name <- gsub(",$", "", imdb250$director_name)
imdb250$writer_id <- gsub(",$", "", imdb250$writer_id)
imdb250$writer_name <- gsub(",$", "", imdb250$writer_name)
```

```{r}
# remove "Writers," from writer_name column
imdb250$writer_name <- gsub("Writers,", "", imdb250$writer_name)
```

```{r}
# remove commas from imdb_votes column
imdb250$imdb_votes <- gsub(",", "", imdb250$imdb_votes)

# change column type
imdb250$imdb_votes <- as.integer(imdb250$imdb_votes)
```

Change duration column to integer column
```{r}
durationdf <- imdb250 %>% select(duration)

durationdf <- durationdf %>%
  separate_wider_delim(duration, delim = " ", names_sep = "", too_few = "align_start")

# the movie in row 195 is only 45 minutes but the value is in the hour column, move to minute column
durationdf[195,'duration2'] <- durationdf[195,'duration1']
durationdf[195, 'duration1'] <- "0"

durationdf$duration1 <- gsub("h", "", durationdf$duration1)
durationdf$duration2 <- gsub("m", "", durationdf$duration2)

durationdf$duration1 <- as.integer(durationdf$duration1)
durationdf$duration2 <- as.integer(durationdf$duration2)

durationdf$duration1 <- 60*(durationdf$duration1)

imdb250$duration <- rowSums(durationdf[,c('duration1', 'duration2')], na.rm = TRUE)
```

```{r}
# re-print head to check for changes
head(imdb250)
```


### Summarize Data
```{r}
summary(imdb250)
```
The only numeric columns are imdb_rank, year, and imdb_rating. All other columns are strings.

### Identify Missing Values
```{r}
sapply(imdb250, function(x) sum(is.na(x)))
```
There are no missing values in the data frame. We can start preparing the data to be loaded in as tables.

# Prepare Data to Fit Database Schema

## Format Data for "movies" Table
```{r}
# create a data frame that contains the columns the belong in the movies table
moviesdf <- imdb250 %>% select(movie_id, title, link, year, imdb_rank, imdb_votes, imdb_rating, mpaa_rating, duration, plot)
```

## Format Data for "genres" table
```{r}
# get unique genres
allGenres <- imdb250$genre
splitGenres <- c()
for(k in 1:length(allGenres)) {
  splitGenres = append(splitGenres, unlist(strsplit(allGenres[[k]], ",")))
}
genres <- list(unique(splitGenres))
genresID <- c(0:20)
genresdf <- data.frame(genresID, genres)
colnames(genresdf) <- c('genre_id', 'genre')
```

## Format Data for "movie_genres" table

```{r}
# get paired movie_id and genre
movieGenresdf <- imdb250 %>% select(movie_id, genre)

# un-nest writer names
movieGenresdf <- movieGenresdf %>%
  separate_longer_delim(genre, delim = ',')

# replace names with IDs from genresdf
genres_to_ids <- movieGenresdf[, 'genre']

row_count <- nrow(genresdf)

for (i in 1:row_count) {
  search_value <- genresdf[i, 2]
  replace_value <- genresdf[i, 1]
  
  genres_to_ids <- sapply(genres_to_ids, function(x) {
    x[x==search_value] <- replace_value
    return(x)
  })
}

movieGenresdf[,'genre_id'] <- genres_to_ids

# drop genre column
movieGenresdf <- movieGenresdf %>% select(-c(genre))
```


## Format Data for "actors" table
```{r}
# get unique actor names
allActors <- imdb250$cast_name
splitActors <- c()
for(k in 1:length(allActors)) {
  splitActors = append(splitActors, unlist(strsplit(allActors[[k]], ",")))
}
actors <- list(unique(splitActors))
actorIDs <- c(0:16332)

# save to data frame
actorsdf <- data.frame(actorIDs, actors)

# change column name
colnames(actorsdf) <- c('actor_id', 'name')
```

## Format Data for "movie_actors" table
```{r}
# get paired movie_id and actor_name
movieActorsdf <- imdb250 %>% select(movie_id, cast_name)

# un-nest writer names
movieActorsdf <- movieActorsdf %>%
  separate_longer_delim(cast_name, delim = ',')

# replace names with IDs from actorsdf
names_to_ids <- movieActorsdf[, 'cast_name']

row_count <- nrow(actorsdf)

for (i in 1:row_count) {
  search_value <- actorsdf[i, 2]
  replace_value <- actorsdf[i, 1]
  
  names_to_ids <- sapply(names_to_ids, function(x) {
    x[x==search_value] <- replace_value
    return(x)
  })
}

movieActorsdf[,'actor_id'] <- names_to_ids

# drop writer_name column
movieActorsdf <- movieActorsdf %>% select(-c(cast_name))
```

## Format Data for "directors" table
```{r}
# get unique director names
allDirectors <- imdb250$director_name
splitDirectors <- c()
for(k in 1:length(allDirectors)) {
  splitDirectors = append(splitDirectors, unlist(strsplit(allDirectors[[k]], ",")))
}

directors <- list(unique(splitDirectors))
directorIDs <- c(0:176)

directorsdf <- data.frame(directorIDs, directors)

# change column name
colnames(directorsdf) <- c('director_id', 'name')
```

## Format Data for "movie_directors" table
```{r}
# get paired movie_id and director_name
movieDirectorsdf <- imdb250 %>% select(movie_id, director_name)

# un-nest writer names
movieDirectorsdf <- movieDirectorsdf %>%
  separate_longer_delim(director_name, delim = ',')

# replace names with IDs from directorsdf
names_to_ids <- movieDirectorsdf[, 'director_name']

row_count <- nrow(directorsdf)

for (i in 1:row_count) {
  search_value <- directorsdf[i, 2]
  replace_value <- directorsdf[i, 1]
  
  names_to_ids <- sapply(names_to_ids, function(x) {
    x[x==search_value] <- replace_value
    return(x)
  })
}

movieDirectorsdf[,'director_id'] <- names_to_ids

# drop director_name column
movieDirectorsdf <- movieDirectorsdf %>% select(-c(director_name))
```

## Format Data for "writers" table
```{r}
# get unique writer names
allWriters <- imdb250$writer_name
splitWriters <- c()
for(k in 1:length(allWriters)) {
  splitWriters = append(splitWriters, unlist(strsplit(allWriters[[k]], ",")))
}
writers <- list(unique(splitWriters))

# create ID column
writerIDs <- c(0:435)

# create data frame
writersdf <- data.frame(writer_id = writerIDs, name = writers)

# change column name
colnames(writersdf) <- c('writer_id', 'name')
```

## Format Data for "movie_writers" table
```{r}
# get paired movie_id and writer_name
movieWritersdf <- imdb250 %>% select(movie_id, writer_name)

# un-nest writer names
movieWritersdf <- movieWritersdf %>%
  separate_longer_delim(writer_name, delim = ',')

# replace names with IDs from writersdf
names_to_ids <- movieWritersdf[, 'writer_name']

row_count <- nrow(writersdf)

for (i in 1:row_count) {
  search_value <- writersdf[i, 2]
  replace_value <- writersdf[i, 1]
  
  names_to_ids <- sapply(names_to_ids, function(x) {
    x[x==search_value] <- replace_value
    return(x)
  })
}

movieWritersdf[,'writer_id'] <- names_to_ids

# drop writer_name column
movieWritersdf <- movieWritersdf %>% select(-c(writer_name))
```

## Save all dataframes as CSV files
```{r}
write.csv(moviesdf, file = "movies.csv")
write.csv(genresdf, file = "genres.csv")
write.csv(movieGenresdf, file = "movie_genres.csv")
write.csv(actorsdf, file = "actors.csv")
write.csv(movieActorsdf, file = "movie_actors.csv")
write.csv(directorsdf, file = "directors.csv")
write.csv(movieDirectorsdf, file = "movie_directors.csv")
write.csv(writersdf, file = "writers.csv")
write.csv(movieWritersdf, file = "movie_writers.csv")
```

# Create Database and Add Tables

## Create Connection and Database
```{r}
conn <- dbConnect(RSQLite::SQLite(), dbname = "IMDbTop250.db")
```

## Write Data to tables

movies Table
```{r eval=FALSE}
# movie table
dbWriteTable(conn, "movies", value = moviesdf, append = TRUE)
```

genres Table
```{r eval=FALSE}
# genres table
dbWriteTable(conn, "genres", value = genresdf, append = TRUE)
```

movie_genres Table
```{r eval=FALSE}
# movie_genres table
dbWriteTable(conn, "movie_genres", value = movieGenresdf, append = TRUE)
```

actors Table
```{r eval=FALSE}
# actors table
dbWriteTable(conn, "actors", value = actorsdf, append = TRUE)
```

movie_actors Table
```{r eval=FALSE}
# movie_actors table
dbWriteTable(conn, "movie_actors", value = movieActorsdf, append = TRUE)
```

directors Table
```{r eval=FALSE}
# directors table
dbWriteTable(conn, "directors", value = directorsdf)
```

movie_directors Table
```{r eval=FALSE}
# movie_directors table
dbWriteTable(conn, "movie_directors", value = movieDirectorsdf)
```

writers table
```{r eval=FALSE}
# writers table
dbWriteTable(conn, "writers", value = writersdf)
```

movie_writers table
```{r eval=FALSE}
# movie_writers table
dbWriteTable(conn, "movie_writers", value = movieWritersdf)
```


### Check that tables were added to the database
```{r eval=FALSE}
# list all tables
dbListTables(conn)
```

# Query the Database

## Query 1: List the first 10 movies
```{r}
dbGetQuery(conn, "SELECT * FROM movies LIMIT 10")
```

## Query 2: Get the average IMDb rating of the movies in the database
```{r}
dbGetQuery(conn, "SELECT AVG(imdb_rating) AS 'average_rating' FROM movies")
```

## Query 3: Return the title, release year, rank, IMDB rating, MPAA rating, and plot of movies directed by David Fincher
```{r}
dbGetQuery(conn, "SELECT title, directors.name AS 'director', year, imdb_rank, imdb_rating, mpaa_rating, duration, plot 
                  FROM movies 
                  INNER JOIN directors ON movie_directors.director_id = directors.director_id
                  INNER JOIN movie_directors ON movies.movie_id = movie_directors.movie_id
                  WHERE name = 'David Fincher'
                  ORDER BY movies.imdb_rank")
```

## Query 4: Return the 10 writers with the most writing credits
```{r}
dbGetQuery(conn, "SELECT writers.writer_id, name, COUNT(*) AS 'number_of_credits'
                  FROM writers
                  INNER JOIN movie_writers ON writers.writer_id = movie_writers.writer_id
                  GROUP BY name
                  HAVING COUNT(*) > 1
                  ORDER BY number_of_credits DESC
                  LIMIT 10")
```


## Disconnect From Database
```{r}
dbDisconnect(conn)
```


# Sources
https://medium.com/veri-analizi/replace-values-in-dataframe-from-another-dataframe-204b97700ba7


